The role of cognateness in non-native spoken word recognition

### 2024-05-08 manuscript/manuscript.qmd
New analyses for Experiment 3

---
title: "The role of cognateness in non-native spoken word recognition"

abstract: |
    Understanding spoken words in a non-native language is easier when the words are phonologically similar to their translations into the listener's native language are phonologically similar. Phonological similarity plays a more important role in the language performance of low-proficiency bilinguals, compared to that of high-proficiency bilinguals., and This suggests that low-proficiency bilinguals may points to the latter as relying more strongly on cross-language word-to-word links (lexical route), while high-proficiency bilinguals rely more on word-concept associations (conceptual route) than on cross-language word-to-word links (lexical route). Disentangling bilinguals’ reliance on the conceptual and lexical and conceptual routes during translation is challenging, given that both sources of information are present during spoken word comprehensionto some extent in both high- and low-proficiency bilinguals. In this study, we tested English and Spanish native speakers in a translation elicitation task, in which they listened to words in an unfamiliar language (Spanish or Catalan for English natives, and Catalan for Spanish natives), and then had to translate them to their native language. Given their lack of previous knowledge ion the unfamiliar tested language, phonological similarity between the presented words and their correct translations was the only cue available for participants to succeed in the task. This allowed us to explore the informativeness of the lexical route for word translation, in the absence of information from the conceptual route. Our results indicate that participants benefited from the phonological similarity between translation equivalents: the more similar the word they listened to and its translational equivalent, the higher the probability of a correct translation. These results point to cross-language word-to-word links as providing rich information during word translation.
---

```{r}
#| label: setup
#| message: false
#| warning: false
#| include: false

library(targets)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gt)
library(gtExtras)
library(brms)
library(tidybayes)

@@

# import objects
tar_config_set(store = here::here("_targets"),
                        script = here::here("_targets.R"))
tar_load_globals()
tar_load(participants)
tar_load(exp_participants)
tar_load(quest_participants)
tar_load(stimuli)

theme_set(theme_ggdist())

@@

Listening to non-native speech is costlier than listening to native speech, even for highly proficient bilinguals, and especially in acoustically adverse situations [@lecumberri2010non; @takata1990english]. However, there are some sources of information which may make non-native speech comprehension easier. One of the sources of this increased difficulty stems from the possible mismatch between the phonology of the native and the non-native languages: some acoustic features embedded in the non-native speech signal do not overlap perfectly with any phonemic category in the listener’s native language. For example, imagine a Spanish native, with no previous familiarity with any other language, who listens for the first time to French may encounter the word /`r knitr::asis_output("\U0070\U0254\U0281\U0074")`/ (*porte*), translation of door in French. The voiced uvular fricative consonant /`r knitr::asis_output("\U0281")`/ and the open-mid back rounded vowel /`r knitr::asis_output("\U0254")`/ do not exist in Spanish as phonemic categories. These discrepancies can nonetheless be perceived as allophonic variations the native (Spanish) phonemes /r/ and /o/ [@best2001discrimination], and the non-native speech signal may engage lexical processing mechanisms [@weber2004lexical], leading to the activation and selection, and therefore the recognition of its translation to Spanish /`r knitr::asis_output("\U02C8\U0070\U0077\U0065\U0072\U002E\U0074\U0061")`/ (*puerta*). The recognition of this French, unfamiliar word relies almost entirely on its similarity with its Spanish translation. Conversely, imagine now that the same Spanish listener hears English for the first time and encounters the word /`r knitr::asis_output("\U0064\U0254\U02D0")`/ (*door*). This time, phonological similarity is of little help: /`r knitr::asis_output("\U02C8\U0070\U0077\U0065\U0072\U002E\U0074\U0061")`/ and /`r knitr::asis_output("\U0064\U0254\U02D0")`/ share very few phonemes. It would come as a surprise if the listener was able to translate the English word successfully, if not by relying on previously learnt word-concept associations (which they lack, as they are unfamiliar to English).

@@

Early theoretical accounts of this 'Cognate effect' proposed that the impact of cognateness on lexical processing depends on bilinguals’ proficiency in their second language (L2) [@potter1984lexical; see @chen1989patterns for similar results]. The reasoning behind this claim was the assumption that second-language learners start acquiring words in their L2 by associating them to their translation in their L1 (lexical route), instead of directly to their shared concept (conceptual route). This word-to-word connectivity should be sensitive to form-similarity (i.e., cognateness): the more similar the two word- forms are, the stronger the connection. As learners become more proficient, the connection between L2 representations and their meaning grows stronger, and L2 word processing becomes less reliant on word-word connections between L1 and L2 representations. Cognates subsequently exert less impact on lexical processing in proficient speakers [see @andras2022cognate for recent experimental evidence for this claim]. The Revised Hierarchical Model [RHM, @kroll1994category] captured this hypothesis and predicted that translating words from L1 to L2 should take longer than translating from L2 to L1. The rationale behind this prediction was that L1-to-L2 translation relies more strongly on word-to-word links between L1 and L2 representations, whereas L2-to-L1 translation relies more strongly on the mediation between the concept and the two word- forms. L1-to-L2 translation should therefore be more sensitive to the form similarity between the L1 and the L2 representations: cognate words should be retrieved faster than non-cognate words during L1-to-L2 translation.

To test these predictions, @degroot1994forward and @de1992determinants asked Dutch natives with high (but non-native) English proficiency to translate Dutch words to English (L1-to-L2) or English words to Dutch (L2-to-L1). Although participants’ performance in both conditions was roughly equivalent for cognates, non-cognates were translated more slowly from Dutch to English than from English to Dutch, suggesting that L1-to-L2 translation is more sensitive to form-similarity than L2-to-L1 translation. Conversely, participants’ performance was more sensitive to semantic variables (e.g., concreteness) when translating English words to Dutch, compared to when translating Dutch words to English, suggesting that L2-to-L1 translation was more sensitive to the shared conceptual properties of the translation pair.

@@

Since the RHM model was proposed, later studies on monolingual populations have focused how on the role of the network of connections that words establish networks of connections with each other at the form-level (phonological or orthographic) or at the conceptual level. For instance, the Neighbourhood Activation Model [NAM, @luce1998recognizing] highlighted how lexical selection is affected by the number of phonological neighbours that the selected word has. Words surrounded by a larger number of phonological neighbours (words that only differ in one phoneme from the target word) are responded to more slowly and less accurately than words from sparser phonological neighbourhoods [@goldinger1989priming; @luce1990similarity], especially when these neighbours’ lexical frequency is higher [@luce1998recognizing].

The existence of phonological links between words is supported with experimental evidence from the visual world paradigm and phonological priming. @meyer2007activation showed that target pictures were named faster when presented together with a phonologically-related distractor, suggesting that the activation of shared phonological information allowed the target to be activated and produced faster. There are also findings suggesting that facilitatory or inhibitory effects on target recognition depend on the degree of similarity the competitor shares with the target [@hamburger1996phonological; @dufour2003lexical]. When there is only mismatch of one phoneme between prime and target, inhibitory effects are exerted on target recognition [@dufour2003lexical]. This inhibition was proposed to be due to strong lexical activation of the prime which results in lexical interference, as opposed to simple phonological activation from less similar primes that triggers facilitation.

These findings of phonological links between words in the monolingual lexicon pose an important question in the field of bilingualism research: do word representations in one language form part of the phonological or orthographic neighbours of the other language? @weber2004lexical observed that bilinguals displayed lexical competition effects from words in their native language when performing a word-referent matching task in their L2. This highlights that lexical competition for bilinguals during language comprehension may not only come from the tested language but also the native language, despite the native language not being actively required for the task. Other studies have similarly found interference when spoken target words in one language had phonological overlap with a word in the participant’s other language, even when the other language was not required by the task [@spivey1999cross; @desroches2022dynamics]. Such patterns have even been found with young children less than 4 years old [@von2012language].

The Bilingual Interactive Activation (BIA/BIA+) [@van1998orthographic; @dijkstra2002architecture] addressed this issue and suggested that lexical representations from both languages establish both excitatory and inhibitory connections with each other, resulting in an integrated lexicon for both languages. This is, in line with connectionist approaches to lexical processing [@mcclelland1981interactive]. Evidence supporting such an integrated lexicon is provided by the fact that bilinguals’ performance in word recognition tasks is sensitive to the orthographic neighbourhood density of the presented word’s translation. When bilingual participants are presented with words in one language in a lexical decision task, translating them to the other language took longer when they were part of large neighbourhoods [@van1998orthographic]. This suggests that presented words activated orthographic neighbourhoods in the other language, which competed for selection with the target word during recognition.

@@

In the present study, we tested listeners’ reliance on phonological similarity when translating words from an unfamiliar language to their native language, in order to investigate the plausibility of word-to-word connections during L1-to-L2 translation. We tested English native speakers and Spanish native speakers in a translation elicitation task. This paradigm has been used by bilingualism researchers to identify easily-recognisable cognates for experimental studies. Using this paradigm, we aim to delve deeper into the factors affecting foreign word recognition by monolingual participants with no prior exposure to the language.

@@

Following @luce1998recognizing's NAM model, we further hypothesised that if participants are able to correctly translate words from an unfamiliar language based on their phonological similarity to the target word, similar sounding but incorrect translations (e.g., false friends) should be activated too. If this is the case, non-native words that have more phonological neighbours should be less likely to be correctly translated, especially if such neighbours have higher lexical frequency and are phonologically closer to the presented word than the target translation. For instance, one could expect English participants to incorrectly translate the Spanish word *botón* as bottom instead of as its correct translation *button* (because the lexical frequency of bottom is higher than that of button and also phonological closer). This combined constraint led us to calculate a specific form of phonological neighbours where a neighbour is counted only if it is higher frequency and has stronger phonological similarity than the target translation as judged by Levenshtein distance. If the phonological neighbourhood density of the target translation affects participants’ performance negatively, this would suggest that non-native words trigger competition between words in the native language during translation.

We tested three groups of participants in this translation elicitation task: English natives translating Spanish words, English natives translating Catalan words, and Spanish natives translating Catalan words. Catalan and Spanish are both Romance languages whose close typological distance is reflected in the fact that they share many cognates, whereas English is a Germanic language that shares considerably fewer cognates with Catalan and Spanish. By testing participants translating words from typologically close and distant languages, we expected to widen the range of the phonological similarity scores of the translation pairs involved in the experimental task, therefore allowing us to explore potential cross-language differences in participant's performance.

Following @luce1998recognizing's NAM model, we further hypothesised that if participants are able to correctly translate words from an unfamiliar language based on their phonological similarity to the target word, similar sounding but incorrect translations (e.g., false friends) should be activated too. If this is the case,  has more phonological neighbours should be less likely to be correctly translated, especially if such neighbours have higher lexical frequency than the target translation. For instance, one could expect English participants to incorrectly translate the Spanish word *botón* as *bottom* instead of as its correct translation *button* (because the lexical frequency of *bottom* is higher than that of *button*). If the phonological neighbourhood density of the target translation affects participants' performance negatively, this would suggest that non-native words trigger competition between words in the native language during translation.

@@

```{r}
#| label: participants
tbl_data <- participants |> 
    mutate(exp = case_when(
        source=="Experiment" & group!= "cat-SPA" ~ "Experiment 1",
        source=="Experiment" & group== "cat-SPA" ~ "Experiment 2",
        source=="Questionnaire" ~ "Experiment 3",
    )) |> 
    summarise(n = n(),
              n_excluded = sum(!valid_participant),
              across(age, lst(mean, sd, min, max)),
              knows_l2 = sum(l_2 != "None"),
              l2_lst = lst(unique(l_2)[unique(l_2)!="None"]),
              .by = c(group, exp))

@@

```{r}
#| label: participants-excluded-1
excluded_lang <- participants_1 |> 
    filter(l_2_writ_prod > 3 |
               l_2_oral_comp > 3 |
               cat_oral_comp > 3 | 
               cat_writ_prod > 3 |
               spa_oral_comp > 3 |
               spa_writ_prod > 3) |> 
    nrow()

excluded_impairment <- participants_1 |> 
    filter(has_language_problems) |> 
    nrow()

excluded_ntrials <- participants_1 |> 
    filter(invalid_participant_trials) |> 
    nrow()
```
Participants were recruited via Prolific (5£ compensation) and SONA (compensation in academic credits), and gave informed consent before providing any data and the study was conducted in accordance with ethical standards of the Declaration of Helsinki and the protocol was by the University of Oxford Medical Sciences Inter-Divisional Research Ethics Committee (IDREC) (R60939/RE007). Participants were asked to complete the experiment using a laptop in a quiet place with good internet connection. We excluded data from participants that (a) self-rated their oral and/or written skills in Catalan and Spanish, or any other second language as four or higher in a five-point scale (*n* = `r excluded_lang`), (b) were diagnosed with a language (*n* = `r excluded_impairment`), or (c) did not contribute more than 80% of valid trials (*n* = `r excluded_ntrials`).

@@

```{r}
stim_exc_freq <- stimuli |> 
    mutate(is_na = is.na(freq_zipf_2)) |> 
    count(group, is_na) |> 
    filter(is_na)
stim_exc_freq <- split(stim_exc_freq, stim_exc_freq$group)
names(stim_exc_freq) <- janitor::make_clean_names(names(stim_exc_freq))
```

@@

Each trial started with a yellow fixation point presented during one second on the centre of the screen over a black background. After one second, the audio started playing while the dot remained being displayed until the audio ended. Upon the offset of the fixation point and audio, participants were prompted to write their answer by a ">" symbol. Typed letters were displayed in the screen in real time to provide visual feed-back to participants. Participants were allowed to correct their answer. Then, participants pressed the RETURN/ENTER key to confirm their answer and start and new trial. We excluded trials where the participant's answer was not a valid word in the target language or no answer was given. Trials in which the response was mistyped by only one character were counted as correct (e.g., "pengiun" instead of "penguin"), as long as the response did not correspond to a distinct word. Trials in which participants took longer than 10 seconds to respond were also excluded. Participants contributed a total of `r nrow(exp_responses)` valid trials (`r nrow(exp_responses[exp_responses$group %in% c("cat-ENG", "cat-SPA"),])` in Catalan, `r nrow(exp_responses[exp_responses$group %in% c("spa-ENG"),])` in Spanish). The task took approximately 15 minutes to be completed.

```{r}
#| label: procedure-figure
#| echo: false
#| message: false
#| warning: false
#| out-width: "80%"
#| fig-cap: "Schematic representation of a trial in the experimental task."
include_graphics(here("img", "design.png"))
```

@@

```{r}
lengths <- dataset_1 |> 
    mutate(len = nchar(response)) |> 
    summarise(across(len,
                     lst(mean, median, sd, min, max)),
              .by = c(group)) 
lengths <- split(lengths, lengths$group)

@@

#### Predictors

We measured the phonological similarity between presented words and their target translation by computing the Levenshtein similarity between their X-SAMPA transcriptions using the `stringsim` function of the `stringdist` R package [@van2014stringdist]. This function computes the inverse of the Levenshtein distance between two character strings as a proportion. First, it computes the edit distance between two character strings (in this case, phoneme transcriptions) by counting the number of additions, deletions, and substitutions necessary to make both strings identical [@levenshtein1966binary]. This measure is then divided by the maximum distance (according to the length of the longest string) and then subtracted from 1. The result is a percentage score, with 0% indicating no similarity between the two strings, and 100% indicating a perfect match between the two strings. We computed this similarity measure (referred simply as Levenshtein, from here on) for every translation pair in our stimuli lists. For example, the *table* (teɪbəl)-*mesa* (mesa) translation pair had a `r percent(stringsim(enc2utf8("teɪbəl"), enc2utf8("mesa")))` similarity, while the *train* (trEIn)-*tren* (t`r knitr::asis_output("\U027E")`En) translation pair had a `r percent(stringsim(enc2utf8("trEIn"), enc2utf8("tɾEn")))` similarity. Figure 1 summarises the lexical frequency, phonological neighbourhood density and phonological overlap of the words included in the Catalan and the Spanish lists.

>>>>

---
title: "The role of cognateness in non-native spoken word recognition"
---

```{r}
#| label: setup
library(targets)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(gt)
library(gtExtras)
library(brms)
library(tidybayes)

@@

# import objects
tar_config_set(store = here::here("_targets"),
               script = here::here("_targets.R"))
tar_load_globals()
tar_load(participants)
tar_load(exp_participants)
tar_load(exp_responses)
tar_load(quest_responses)
tar_load(quest_participants)
tar_load(stimuli)
tar_load(epreds)
tar_load(dataset_1)
tar_load(dataset_2)
tar_load(dataset_3)
tar_load(exp_1_m0)
tar_load(exp_2_m0)
tar_load(exp_3_m1)

theme_set(theme_ggdist() + 
              theme(panel.grid.major.y = element_line(colour = "grey",
                                                      linetype = "dotted")))

@@

Listening to non-native speech is costlier than listening to native speech, even for highly proficient bilinguals, and especially in acoustically adverse situations [@lecumberri2010non; @takata1990english]. However, there are some sources of information which may make non-native speech comprehension easier. One of the sources of this increased difficulty stems from the possible mismatch between the phonology of the native and the non-native languages: some acoustic features embedded in the non-native speech signal do not overlap perfectly with any phonemic category in the listener's native language. For example, imagine a Spanish native, with no previous familiarity with any other language, who listens for the first time to French may encounter the word /`r knitr::asis_output("\U0070\U0254\U0281\U0074")`/ (*porte*), translation of door in French. The voiced uvular fricative consonant /`r knitr::asis_output("\U0281")`/ and the open-mid back rounded vowel /`r knitr::asis_output("\U0254")`/ do not exist in Spanish as phonemic categories. These discrepancies can nonetheless be perceived as allophonic variations the native (Spanish) phonemes /r/ and /o/ [@best2001discrimination], and the non-native speech signal may engage lexical processing mechanisms [@weber2004lexical], leading to the activation and selection, and therefore the recognition of its translation to Spanish /`r knitr::asis_output("\U02C8\U0070\U0077\U0065\U0072\U002E\U0074\U0061")`/ (*puerta*). The recognition of this French, unfamiliar word relies almost entirely on its similarity with its Spanish translation. Conversely, imagine now that the same Spanish listener hears English for the first time and encounters the word /`r knitr::asis_output("\U0064\U0254\U02D0")`/ (*door*). This time, phonological similarity is of little help: /`r knitr::asis_output("\U02C8\U0070\U0077\U0065\U0072\U002E\U0074\U0061")`/ and /`r knitr::asis_output("\U0064\U0254\U02D0")`/ share very few phonemes. It would come as a surprise if the listener was able to translate the English word successfully, if not by relying on previously learnt word-concept associations (which they lack, as they are unfamiliar to English).

@@

Early theoretical accounts of this 'Cognate effect' proposed that the impact of cognateness on lexical processing depends on bilinguals' proficiency in their second language (L2) [@potter1984lexical; see @chen1989patterns for similar results]. The reasoning behind this claim was the assumption that second-language learners start acquiring words in their L2 by associating them to their translation in their L1 (lexical route), instead of directly to their shared concept (conceptual route). This word-to-word connectivity should be sensitive to form-similarity (i.e., cognateness): the more similar the two word- forms are, the stronger the connection. As learners become more proficient, the connection between L2 representations and their meaning grows stronger, and L2 word processing becomes less reliant on word-word connections between L1 and L2 representations. Cognates subsequently exert less impact on lexical processing in proficient speakers [see @andras2022cognate for recent experimental evidence for this claim]. The Revised Hierarchical Model [RHM, @kroll1994category] captured this hypothesis and predicted that translating words from L1 to L2 should take longer than translating from L2 to L1. The rationale behind this prediction was that L1-to-L2 translation relies more strongly on word-to-word links between L1 and L2 representations, where L2-to-L1 translation relies more strongly on the mediation between the concept and the two word- forms. L1-to-L2 translation should therefore be more sensitive to the form similarity between the L1 and the L2 representations: cognate words should be retrieved faster than non-cognate words during L1-to-L2 translation.

To test these predictions, @degroot1994forward and @de1992determinants asked Dutch natives with high (but non-native) English proficiency to translate Dutch words to English (L1-to-L2) or English words to Dutch (L2-to-L1). Although participants' performance in both conditions was roughly equivalent for cognates, non-cognates were translated more slowly from Dutch to English than from English to Dutch, suggesting that L1-to-L2 translation is more sensitive to form-similarity than L2-to-L1 translation. Conversely, participants' performance was more sensitive to semantic variables (e.g., concreteness) when translating English words to Dutch, compared to when translating Dutch words to English, suggesting that L2-to-L1 translation was more sensitive to the shared conceptual properties of the translation pair.

@@

Since the RHM model was proposed, later studies on monolingual populations have focused how on the role of the network of connections that words establish networks of connections with each other at the form-level (phonological or orthographic) or at the conceptual level. For instance, the Neighbourhood Activation Model [NAM, @luce1998recognizing] highlighted how lexical selection is affected by the number of phonological neighbours that the selected word has. Words surrounded by a larger number of phonological neighbours (words that only differ in one phoneme from the target word) are responded to more slowly and less accurately than words from sparser phonological neighbourhoods [@goldinger1989priming; @luce1990similarity], especially when these neighbours' lexical frequency is higher [@luce1998recognizing].

The existence of phonological links between words is supported with experimental evidence from the visual world paradigm and phonological priming. @meyer2007activation showed that target pictures were named faster when presented together with a phonologically-related distractor, suggesting that the activation of shared phonological information allowed the target to be activated and produced faster. there are also findings suggesting that facilitatory or inhibitory effects on target recognition depend on the degree of similarity the competitor shares with the target [@hamburger1996phonological; @dufour2003lexical]. When there is only mismatch of one phoneme between prime and target, inhibitory effects are exerted on target recognition [@dufour2003lexical]. This inhibition was proposed to be due to strong lexical activation of the prime which results in lexical interference, as opposed to simple phonological activation from less similar primes that triggers facilitation.

These findings of phonological links between words in the monolingual lexicon pose an important question in the field of bilingualism research: do word representations in one language form part of the phonological or orthographic neighbours of the other language? @weber2004lexical observed that bilinguals displayed lexical competition effects from words in their native language when performing a word-referent matching task in their L2. This highlights that lexical competition for bilinguals during language comprehension may not only come from the tested language but also the native language, despite the native language not being actively required for the task. Other studies have similarly found interference when spoken target words in one language had phonological overlap with a word in the participant's other language, even when the other language was not required by the task [@spivey1999cross; @desroches2022dynamics]. Such patterns have even been found with young children less than 4 years old [@von2012language].

The Bilingual Interactive Activation (BIA/BIA+) [@van1998orthographic; @dijkstra2002architecture] addressed this issue and suggested that lexical representations from both languages establish both excitatory and inhibitory connections with each other, resulting in an integrated lexicon for both languages. This is, in line with connectionist approaches to lexical processing [@mcclelland1981interactive]. Evidence supporting such an integrated lexicon is provided by the fact that bilinguals' performance in word recognition tasks is sensitive to the orthographic neighbourhood density of the presented word's translation. When bilingual participants are presented with words in one language in a lexical decision task, translating them to the other language took longer when they were part of large neighbourhoods [@van1998orthographic]. This suggests that presented words activated orthographic neighbourhoods in the other language, which competed for selection with the target word during recognition.

@@

In the present study, we tested listeners' reliance on phonological similarity when translating words from an unfamiliar language to their native language, in order to investigate the plausibility of word-to-word connections during L1-to-L2 translation. We tested English native speakers and Spanish native speakers in a translation elicitation task. This paradigm has been used by bilingualism researchers to identify easily-recognisable cognates for experimental studies. Using this paradigm, we aim to delve deeper into the factors affecting foreign word recognition by monolingual participants with no prior exposure to the language.

@@

Following @luce1998recognizing's NAM model, we further hypothesised that if participants are able to correctly translate words from an unfamiliar language based on their phonological similarity to the target word, similar sounding but incorrect translations (e.g., false friends) should be activated too. If this is the case, non-native words that have more phonological neighbours should be less likely to be correctly translated, especially if such neighbours have higher lexical frequency and are phonologically closer to the presented word than the target translation. For instance, one could expect English participants to incorrectly translate the Spanish word *botón* as bottom instead of as its correct translation *button* (because the lexical frequency of *bottom* is higher than that of *button* and also phonological closer). This combined constraint led us to calculate a specific form of phonological neighbours where a neighbour is counted only if it is higher frequency and has stronger phonological similarity than the target translation as judged by Levenshtein distance. If the phonological neighbourhood density of the target translation affects participants' performance negatively, this would suggest that non-native words trigger competition between words in the native language during translation.

In Experiment 1, we collected data from two groups of British English natives. One group was presented with Catalan words, the other with Spanish words. We examined the extent to which participants were able to use the phonological similarity between the presented (Catalan or Spanish) word and its translation (i.e., cognateness) to provide accurate responses. In Experiment 2, we tested a group of (European) Spanish natives in the same task, who were presented with Catalan words. Catalan and Spanish are both Romance languages whose close typological distance is reflected in the fact that they share many cognates, where English is a Germanic language that shares considerably fewer cognates with Catalan and Spanish. By testing participants translating words from typologically close and distant languages, we expected to widen the range of the phonological similarity scores of the translation pairs involved in the experimental task, therefore allowing us to explore potential cross-language differences in participant's performance. Participants in Experiments 1 and 2 were surprisingly good at translation some words, which had low phonological similarity with their correct translation. In Experiment 3, we collected additional data from a new group of British English natives, who were presented with Catalan and Spanish words. In contrast with the procedure in Experiment 1, participants in Experiment 3 where asked, after providing a response in each trial, to report whether they had previous knowledge of the meaning of the presented word (e.g., some Spanish words are common knowledge, even for monolingual speakers of English). Finally, we present analyses on the joint data sets of Experiments 1 to 3.

@@

Participants were recruited via Prolific (5£ compensation) and SONA (compensation in academic credits), and gave informed consent before providing any data and the study was conducted in accordance with ethical standards of the Declaration of Helsinki and the protocol was by the University of Oxford Medical Sciences Inter-Divisional Research Ethics Committee (IDREC) (R60939/RE007). Participants were asked to complete the experiment using a laptop in a quiet place with good internet connection.  Data collection took place from `r collection_dates[1]` to `r collection_dates[2]`. We collected data from `r n_participants` British English native participants living in United Kingdom (*Mean* = `r round(participants_age$age_mean, 2)` years, *SD* = `r round(participants_age$age_sd, 2)`, Range = `r round(participants_age$age_min, 2)`-`r round(participants_age$age_max, 2)`, `r n_female` female).

```{r tbl-participants}
#| label: tbl-participants
tbl_data <- participants |> 
    mutate(exp = case_when(
        source=="Experiment" & group!= "cat-SPA" ~ "Experiment 1",
        source=="Experiment" & group== "cat-SPA" ~ "Experiment 2",
        source=="Questionnaire" ~ "Experiment 3",
    )) |> 
    summarise(n = n(),
              n_excluded = sum(!valid_status=="Valid"),
              across(age, lst(mean, sd, min, max)),
              knows_l2 = sum(l2 != "None"),
              l2_lst = lst(unique(l2)[unique(l2)!="None"]),
              .by = c(group, exp))

@@

@@

```{r}
stim_exc_freq <- stimuli |> 
    mutate(is_na = is.na(freq_zipf_2)) |> 
    count(group, is_na) |> 
    filter(is_na)
stim_exc_freq <- split(stim_exc_freq, stim_exc_freq$group)
names(stim_exc_freq) <- janitor::make_clean_names(names(stim_exc_freq))

stim_exc_freq_n <- sum(map_int(stim_exc_freq, "n"))
```

For each translation pair, we defined three predictors of interest: the lexical frequency of the correct translation (*Frequency*), the phonological similarity between the presented (Catalan or Spanish) word and their correct English translation (*Cognateness*), and presented word's number of cross-language phonological neighbours (*CLPN*).

We included *Frequency* under the hypothesis that, keeping other predictors constant, participants would translate higher-frequency words more accurately and faster than lower-frequency words. Lexical frequencies of correct translations were extracted from SUBTLEX-UK [@van2014subtlex], and transformed to Zipf scores [@van2014subtlex]. Words in the stimuli list without a lexical frequency value were excluded from data analysis (`r stim_exc_freq$cat_eng$n` in the Catalan list, `r stim_exc_freq$spa_eng$n` in the Spanish list).

We calculated *Cognatenesss*, our main predictor of interest, by computing the Levenshtein similarity between the X-SAMPA transcriptions of each pair of translations using the `stringdist` R package [@van2014stringdist]. The Levenshtein distance computes the edit distance between two character strings (in this case, phoneme transcriptions) by counting the number of additions, deletions, and substitutions necessary to make both strings identical [@levenshtein1966binary]. We divided this edit distance by the number of characters included in the longest X-SAMPA transcription of the translation pair. This results in a proportion score, in which values closer to zero correspond to lower Levenshtein distances between phonological transcriptions (i.e., higher similarity), and values closer to 1 correspond to higher Levenshtein distances (i.e., lower similarity). This transformation also help account for the fact that the Levenshtein distance tends to increase with the length of the transcriptions. For interpretability, we subtracted this proportion from one, so that values closer to one correspond to higher similarity between phonological transcriptions, and values closer to zero correspond to lower similarity between phonological transcriptions. For example, the *table* (teɪbəl)-*mesa* (mesa) translation pair had a `r scales::percent(stringsim(enc2utf8("teɪbəl"), enc2utf8("mesa")))` similarity, while the *train* (trEIn)-*tren* (t`r knitr::asis_output("\U027E")`En) translation pair had a `r scales::percent(stringsim(enc2utf8("trEIn"), enc2utf8("tɾEn")))` similarity. Figure 1 summarises the lexical frequency, phonological neighbourhood density and phonological overlap of the words included in the Catalan and the Spanish lists.

For each Catalan and Spanish word, we calculated the number of *CLPN* by counting the number of English words with same or higher lexical frequency, and whose phonological transcription (in X-SAMPA format) different in up to one phoneme from that of the presented Catalan or Spanish word. Lexical frequencies and phonological transcriptions were extracted from the multilingual database CLEARPOND [@marian2012clearpond]^[[Phonological trnascriptions in CLEARPOND were generated from eSPEAK, [http://espeak.sourceforge.net/](http://espeak.sourceforge.net/)]].

```{r tbl-stimuli}
#| label: tbl-stimuli
stimuli |> 
    summarise(across(c(freq_zipf_2, lv, neigh_n_h),
                     lst(mean, sd, min, max), 
                     na.rm = TRUE),
              .by = c(group)) |> 
    gt(rowname_col = "group") |> 
    tab_spanner("Frequency", matches("freq")) |> 
    tab_spanner("Cognateness", matches("lv")) |> 
    tab_spanner("CLPN", matches("neigh")) |> 
    fmt_number(is.numeric) |> 
    fmt_integer(c(neigh_n_h_min,
                  neigh_n_h_max)) |> 
    cols_merge_uncert(freq_zipf_2_mean, freq_zipf_2_sd) |> 
    cols_merge_uncert(lv_mean, lv_sd) |> 
    cols_merge_uncert(neigh_n_h_mean, neigh_n_h_sd) |> 
    cols_merge_range(freq_zipf_2_min, freq_zipf_2_max) |> 
    cols_merge_range(lv_min, lv_max) |> 
    cols_merge_range(neigh_n_h_min, neigh_n_h_max) |> 
    cols_label(freq_zipf_2_mean = "Mean ± SD",
               freq_zipf_2_sd = "SD",
               freq_zipf_2_min = "Range",
               lv_mean = "Mean ± SD",
               lv_sd = "SD",
               lv_min = "Range",
               neigh_n_h_mean = "Mean ± SD",
               neigh_n_h_sd = "SD",
               neigh_n_h_min = "Range") |> 
    tab_style(cell_text(weight = "bold"),
              cells_column_spanners()) |> 
    tab_style(cell_text(style = "italic"),
              cells_column_labels())

@@

Each trial started with a yellow fixation point presented during one second on the centre of the screen over a black background. After one second, the audio started playing while the dot remained being displayed until the audio ended. Upon the offset of the fixation point and audio, participants were prompted to write their answer by a `>` symbol. Typed letters were displayed in the screen in real time to provide visual feed-back to participants. Participants were allowed to correct their answer. Then, participants pressed the `RETURN/ENTER` key to confirm their answer and start and new trial. We excluded trials where the participant's answer was not a valid word in the target language or no answer was given. Trials in which the response was mistyped by only one character were counted as correct (e.g., "pengiun" instead of "penguin"), as long as the response did not correspond to a distinct word. Trials in which participants took longer than 10 seconds to respond were also excluded. Participants contributed a total of `r nrow(exp_responses)` valid trials (`r nrow(exp_responses[exp_responses$group %in% c("cat-ENG", "cat-SPA"),])` in Catalan, `r nrow(exp_responses[exp_responses$group %in% c("spa-ENG"),])` in Spanish). The task took approximately 15 minutes to be completed.

```{r}
#| label: procedure-figure
#| echo: false
#| message: false
#| warning: false
#| out-width: "80%"
#| fig-cap: "Schematic representation of a trial in the experimental task."
knitr::include_graphics(here::here("img", "design.png"))
```

@@

We modelled the probability of participants guessing the correct translation of each presented word using a generalised multilevel Bayesian regression model with a Bernoulli logit link distribution. We included as fixed effects the intercept, the main effects of *Frequency*, *Cognateness*, and *CLPN*, and the two-way interaction between *Cognateness* and *CLPN*. We also included participant-level random intercepts and slopes for the the main effects and the interaction. Eq. 1 shows a formal description of the model.

![Eq. 1](_assets/img/model.PNG)

To test the practical relevance of each predictor we followed [@kruschke2018bayesian]. We first specified a region of practical equivalence (*ROPE*) around zero ([-0.1, +0.1], in the logit scale). This area indicates the values of the regression coefficients that we considered as equivalent to zero. We then computed the 95% posterior credible intervals (CrI) of the regression coefficients of interest. Finally, we calculated the proportion of the 95% CrI that fell inside the ROPE. This proportion indicates the probability that the true value of the coefficient is equivalent to zero.

All analyses were performed in R environment [@rcore2019r]. We used the tidyverse family of R packages [@wickham2019tidyverse] to process data and to generate figures. We used the `brms` R package [@burkner2017brms] using the `cmdstanr` back-end to the Stan probabilistic language [@carpenter2017stan] to estimate and compare the models (see Appendix 1 for mode details on the models).

```{r dataset-1}
#| label: dataset-1
dataset <- exp_responses |> 
    filter(group!="cat-SPA") |> 
    group_split(group) |> 
    set_names(c("spa-ENG", "cat-ENG")) |> 
    janitor::clean_names()

dataset <- split(exp_responses, exp_responses$group)
names(dataset) <- janitor::make_clean_names(names(dataset))
r_validity <- map(dataset, 
                  \(x) janitor::clean_names(table(x$response_type)))
n_participants <- map(dataset, \(x) n_distinct(x$participant_id))

excluded_lang <- participants_1 |> 
    filter(l2_writ_prod > 3 |
               l2_oral_comp > 3 |
               cat_oral_comp > 3 | 
               cat_writ_prod > 3 |
               spa_oral_comp > 3 |
               spa_writ_prod > 3) |> 
    nrow()

excluded_impairment <- participants_1 |> 
    filter(valid_status=="Language impairment") |> 
    nrow()

excluded_ntrials <- participants_1 |> 
    filter(valid_status=="Insufficient trials") |> 
    nrow()

dataset_clean <- exp_responses |> 
    filter(group!="cat-SPA") |> 
    dplyr::filter(valid_participant, valid_response) |> 
    group_split(group) |> 
    set_names(c("spa-ENG", "cat-ENG")) |> 
    janitor::clean_names()

lengths <- dataset_1 |> 
    mutate(len = nchar(response)) |> 
    summarise(across(len,
                     lst(mean, median, sd, min, max)),
              .by = c(group)) 
lengths <- split(lengths, lengths$group)

@@

```{r}
c1 <- gather_draws(exp_1_m0, `b_.*`, `sd_.*`, regex = TRUE) |> 
    summarise(across(.value, lst(.median = median, 
                                 .lower = \(x) hdi(x)[1],
                                 .upper = \(x) hdi(x)[2],
                                 .rope = \(x) mean(between(x, -0.1, 0.1))),
                     .names = "{fn}"))
c1 <- split(c1, c1$.variable)
names(c1) <- janitor::make_clean_names(names(c1)) 
```
